---
title: "NYPD SHOOTINGS DATA ANALYSIS"
output:
  html_document: default
  pdf_document: default
date: "2023-06-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###IMPORT DATASETS

```{r get_nypd, message = FALSE}
url_in <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
url_pop <- "https://data.cityofnewyork.us/api/views/xywu-7bv9/rows.csv?accessType=DOWNLOAD"
```

###IMPORT LIBRARY

```{r import_data, message = FALSE}
library(readr)
library(lubridate)
library(dplyr)
library(ggplot2)
library(stringr)
library(tidyverse)
nypd_shooting_main <- read_csv(url_in, show_col_types = FALSE)
nypd_shooting_main
ny_population <- read_csv(url_pop, show_col_types = FALSE)
ny_population
```



###CLEAN DATA
```{r clean_data, message = FALSE}

nypd_shooting <- nypd_shooting_main %>%
  rename(DATE = OCCUR_DATE,
         TIME = OCCUR_TIME)
nypd_shooting <- nypd_shooting %>% 
  mutate(across(-TIME, ~ifelse(is.na(.), "N/A", .)))
nypd_shooting$DATE <- as.Date(nypd_shooting$DATE, format = "%m/%d/%Y")
nypd_shooting <- subset(nypd_shooting, select = c("DATE","TIME", "BORO", "LOC_OF_OCCUR_DESC","LOC_CLASSFCTN_DESC","LOCATION_DESC","PERP_AGE_GROUP","PERP_SEX","PERP_RACE","VIC_AGE_GROUP","VIC_SEX","VIC_RACE")) 





```
```{r join_population,  message = FALSE}
ny_population_2020 <- subset(ny_population, select = c("Borough","2020"))
ny_population_2020$Borough = toupper(ny_population_2020$Borough)
ny_population_2020 <- ny_population_2020 %>% 
  rename(BORO = Borough, POPULATION = "2020")

nypd_shooting <- nypd_shooting %>%
  left_join(ny_population_2020, by = c("BORO"))

nypd_shooting %>%
select (DATE, TIME, BORO, POPULATION)
```



##ANALYZE DATA
##Calculate shootings per thousand
```{r shootings_by_boro,  message = FALSE}
shootings_by_boro <- nypd_shooting %>% 
  group_by(BORO) %>% 
  summarise(Total = n()) %>% 
  arrange(desc(Total))

shootings_by_boro <- shootings_by_boro %>% left_join(ny_population_2020, by = c("BORO"))
data_new <-  shootings_by_boro
shootings_by_boro$SHOOTINGS_PER_THOUSAND <- data_new$Total * 1000/ data_new$POPULATION

shootings_by_boro



```



##Visualization of shootings by BORO. Calcuate shootings per thousand population
```{r vis_of_shootings,  message = FALSE}
ggplot(shootings_by_boro, aes(x=BORO, y=Total, fill=BORO)) +
  geom_bar(stat="identity") +
  xlab("BORO") + ylab("Total shootings") +
  ggtitle("NEW YORK SHOOTINGS") +
  theme_minimal()

ggplot(shootings_by_boro, aes(x=BORO, y=SHOOTINGS_PER_THOUSAND, fill=BORO)) +
  geom_bar(stat="identity") +
  xlab("BORO") + ylab("Shootings per thousand") +
  ggtitle("NEW YORK SHOOTINGS PER THOUSAND") +
  theme_minimal()


```
###BRONX boro has the highest shootings per thousand



##Visualization of shootings by YEAR
```{r shootings_by_year,  message = FALSE}
shootings_by_year <- nypd_shooting %>%
mutate(year = year(DATE)) %>%
group_by(year, BORO) %>%
  summarise(Total = n()) %>% 
  arrange((year))

pivot_data <- shootings_by_year %>% pivot_wider(names_from = BORO, values_from = Total)
pivot_data

shootings_by_year




pivot_data %>% 

  ggplot(aes(x = year, y = BRONX)) +
  geom_line(aes(color = "BRONX")) +
  geom_point(aes(color = "BRONX")) +
  geom_line(aes(y = MANHATTAN, color = "MANHATTAN")) +
  geom_point(aes(y = MANHATTAN, color = "MANHATTAN")) +
  geom_line(aes(y = BROOKLYN, color = "BROOKLYN")) +
  geom_point(aes(y = BROOKLYN, color = "BROOKLYN")) +
    geom_line(aes(y = QUEENS, color = "QUEENS")) +
  geom_point(aes(y = QUEENS, color = "QUEENS")) +
    geom_line(aes(y = `STATEN ISLAND`, color = "STATEN ISLAND")) +
  geom_point(aes(y = `STATEN ISLAND`, color = "STATEN ISLAND")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) + 
  labs(title = str_c("SHOOTINGS EACH YEAR"), y=NULL)
```




##LINEAR MODEL
```{r model, message=FALSE}
mod <- lm(POPULATION ~ Total, data = shootings_by_boro)
summary(mod)
shoot_pred <- shootings_by_boro %>% mutate(pred = predict(mod))
ggplot(shoot_pred, aes(x = POPULATION, y = pred)) +
  geom_point() +
  geom_smooth(method = "lm") +
  xlab("POPULATION OF BORO") +
  ylab("TOTAL SHOOTINGS") +
  ggtitle("TOTAL SHOOTINGS VS POPULATION")
```  

##BIAS IDENTIFICATION

```{r bias, message=FALSE}
shootings_by_loc <- nypd_shooting %>%
 filter(LOC_CLASSFCTN_DESC != 'N/A') %>%
group_by(LOC_CLASSFCTN_DESC) %>%
  summarise(Total = n()) 
shootings_by_loc 
```

```{r bias_plot, message=FALSE}
ggplot(shootings_by_loc, aes(x=LOC_CLASSFCTN_DESC, y=Total, fill=LOC_CLASSFCTN_DESC)) +
  geom_bar(stat="identity") +
  xlab("LOCATION") + ylab("Total shootings") +
  ggtitle("SHOOTING LOCATIONS") +
  theme_minimal()
```

* Before conducting this data analysis, I had a personal bias that the most number of shootings occurs either on housing or commerical buildings either due to disagreement, fights, robbery/attack at home etc.

* This data analysis is a real eye opener for me and it clearly shows that highest number of shootings take place on STREETS. This data helps in clearing out my personal bias. This is a real value add and data speaks above any bias.





###session info
```{r echo=FALSE}
sessionInfo(package = NULL)
```