---
title: "Covid 19 Data Analaysis"
output:
  html_document: default
  pdf_document: default
date: "2023-06-25"
---
###This report contains the data analysis and Visualization of following
Analysis and Visualization of US totals
Analysis and Visualization for State of Illinois cases vs deaths
New case vs new deaths in the state of Alaska
Lowest Deaths per cases
Highest Deaths per cases
Linear model relationship between cases per thousand to deaths per thousand

###Question of Interest
###How new cases related to deaths
###Which states are more prone to deaths and what is their background?
###What is the impact on Vaccination? Did deaths started to reduce after vaccination?
###Is population directly related to cases and deaths?


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


###SOURCE OF DATA

##This data is sourced on the basis of study conducted by Johns Hopkins University.This includes separate dataset for Global and another for US.


```{r echo=FALSE, message=FALSE}
library(tidyverse)
library(lubridate)
library(readr)
library(dplyr)
library(ggplot2)
library(stringr)


url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <- c("time_series_covid19_confirmed_global.csv","time_series_covid19_deaths_global.csv","time_series_covid19_confirmed_US.csv", "time_series_covid19_deaths_US.csv")
urls <- str_c(url_in, file_names)
global_cases <- read_csv(urls[1], show_col_types = FALSE)
global_deaths <- read_csv(urls[2], show_col_types = FALSE)
US_cases <- read_csv(urls[3], show_col_types = FALSE)
US_deaths <- read_csv(urls[4], show_col_types = FALSE)
```

```{r echo=FALSE, message=FALSE}
options(readr.show_col_types = FALSE)
global_cases <- global_cases %>%
  pivot_longer(cols = -c(`Province/State`, `Country/Region`, Lat, Long),
               names_to = "date",
               values_to = "cases") %>%
  select(-c(Lat, Long))

global_deaths <- global_deaths %>%
  pivot_longer(cols = -c(`Province/State`, `Country/Region`, Lat, Long),
               names_to = "date",
               values_to = "deaths") %>%
  select(-c(Lat, Long))
  
global <- global_cases %>%
    full_join(global_deaths) %>%
  rename(Country_Region = `Country/Region`,
         Province_State = `Province/State`) %>%
         mutate(date = mdy(date))

global <- global %>% filter(cases > 0)
global <- global %>%
  unite("Combined_Key", 
  c(Province_State, Country_Region), 
  sep = ", ",
  na.rm = TRUE,
  remove = FALSE)

US_cases <- US_cases %>%
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to = "cases") %>%
  select(Admin2:cases) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))



US_deaths <- US_deaths %>%
  pivot_longer(cols = -(UID:Population),
               names_to = "date",
               values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))


US <- US_cases %>% full_join(US_deaths)



uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
uid <- read_csv(uid_lookup_url, show_col_types = FALSE) %>% select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))

global <- global %>%
  left_join(uid, by = c("Province_State","Country_Region")) %>%
  select(-c(UID, FIPS)) %>%
  select(Province_State, Country_Region, date, cases, deaths, Population, Combined_Key)
```

###US_cases

```{r echo=FALSE}
US_cases 
```

###US_deaths

```{r echo=FALSE}
US_deaths
```

###US

```{r echo=FALSE}
US
```

##Global

```{r echo=FALSE}
global
```

###US_totals

```{r echo=FALSE, message=FALSE}
US_totals  <- US_by_state <- US %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths), 
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths * 1000000/ Population) %>%
  select(Country_Region, date, cases, deaths, deaths_per_mill, Population) %>%
  ungroup()
US_totals
```

###Visualization of US_totals

```{r echo=FALSE, message=FALSE}
US_totals %>% 
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) + 
  labs(title = "COVID19 in US", y=NULL)

```

###US totals new cases

```{r echo=FALSE, message=FALSE}
US_totals <- US_totals %>% mutate(new_cases = cases - lag(cases),                                       new_deaths = deaths - lag(deaths))
US_totals %>% 
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = new_deaths, color = "deaths")) +
  geom_point(aes(y = new_deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) + 
  labs(title = "COVID19 in US", y=NULL)

```

###US_by_state

```{r echo=FALSE, message=FALSE}
US_by_state <- US %>%
  group_by(Province_State, Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths), 
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths * 1000000/ Population) %>%
  select(Province_State, Country_Region, date, cases, deaths, deaths_per_mill, Population) %>%
  ungroup()
US_by_state
```

###Visualization of Illinois

```{r echo=FALSE, message=FALSE}
state <- "Illinois"
US_by_state %>%
  filter(cases > 0) %>%
  filter(Province_State == state) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) + 
  labs(title = str_c("COVID19 in US", state), y=NULL)
```

###New cases

```{r echo=FALSE, message=FALSE}
US_by_state <- US_by_state %>% mutate(new_cases = cases - lag(cases), 
                                      new_deaths = deaths - lag(deaths))
US_by_state

```

```{r echo=FALSE, message=FALSE}
state <- "Alaska"
US_by_state %>% 
  filter(cases > 0) %>%
  filter(Province_State == state) %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) + 
  labs(title = str_c("COVID19 in US", state), y=NULL)

```

###Lowest deaths per cases

```{r echo=FALSE, message=FALSE}
US_state_totals <- US_by_state %>%
  group_by(Province_State) %>%
  summarize(deaths = max(deaths), cases = max(cases), population = max(Population), 
            cases_per_thou = 1000* cases / population,
            deaths_per_thou = 1000* deaths / population) %>%
            filter(cases > 0, population > 0)

US_state_totals %>% 
  slice_min(deaths_per_thou, n = 10)
```

###Highest deaths per cases

```{r echo=FALSE, message=FALSE}
US_state_totals <- US_by_state %>%
  group_by(Province_State) %>%
  summarize(deaths = max(deaths), cases = max(cases), population = max(Population), 
            cases_per_thou = 1000* cases / population,
            deaths_per_thou = 1000* deaths / population) %>%
            filter(cases > 0, population > 0)

US_state_totals %>% 
  slice_max(deaths_per_thou, n = 10)
```

###LINEAR MODEL -  cases per thousand to deaths per thousand

```{r echo=FALSE, message=FALSE}
mod <- lm(deaths_per_thou ~ cases_per_thou, data = US_state_totals)
summary(mod)
US_tot_w_pred <- US_state_totals %>% mutate(pred = predict(mod))
US_tot_w_pred %>% ggplot() + 
    geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") +
    geom_point(aes(x = cases_per_thou, y = pred), color = "red")
```



###CONCLUSION

1.There was a rapid increase on cases between beginning of 2020 and beginning of 2021
2.Analysis and Visualization of US New cases and deaths shows, new deaths started to significantly drop beginning 2022 inspite of new cases being reported. This i suspect is due to more people being vaccinated and chances of deaths after vaccination got significantly lower.
3.Analysis and Visualization for State of Illinois cases vs deaths
4.New case vs new deaths in the state of Alaska remain linearly proportional
5.American samoa have lowest Deaths per cases.Other than islands, Hawaii remains as the state with lowest deaths per cases
6.Arizona has the highest deaths per cases
7.Linear model relationship between cases per thousand to deaths per thousand